Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-12-08 analysis/paper/paper.Rmd
typos fixed

Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions. Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks is minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables that can be further analysed by multivariate statistics [@Slice2007]. Landmark-based morphometrics are widely used for archaeological objects with extreme morphological signature or distinctive points that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure of which the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures. In those cases, outline approaches are more suitable to study shape by focusing more on on the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, which assign points along a outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010]. Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982]. EFA is a powerful method to extract geometric information where outlines are considered as periodic functions and decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely. The use of EFA in archaeology are mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is more suitable for shapes lacking of obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, including the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels in varied forms from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic forms. EFA enables the study of the level of production standardisation to answer questions about the evolution of shape over space and time.

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of a society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator for social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.

>>>>

Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions. Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables that can be further analysed by multivariate statistics [@Slice2007]. Landmark-based morphometrics are widely used for archaeological objects with extreme morphological signature or distinctive points that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure of which the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures. In those cases, outline approaches are more suitable to study shape by focusing more on the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010]. Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982]. EFA is a powerful method to extract geometric information where outlines are considered as periodic functions and decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely. The use of EFA in archaeology is mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is more suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, including the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels in varied forms from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic forms. EFA enables the study of the level of production standardisation to answer questions about the evolution of shape over space and time.

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator of social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in forms.
